---
title: "Gene Expression Dataset Output 🧬"
format:
    html:
        toc: true
        toc-title: "Contents"
        embed-resources: true
        page-layout: full
        theme: united
        code-tools: false
        code-line-numbers: true
        title-block-banner: "#4A7AB5"
knitr:
  opts_chunk:
    collapse: true
    message: false
    warning: false
---

```{=html}
<style>
  body { background-color: #AFC4DE; }
</style>
```
::: {.column-screen-inset-left}
This document contains all the results from and code used to complete the 'Gene Expression' portion of the LIFE4138 Coursework.
:::

# Importing the data 📊 {.column-screen-inset-left}
::: {.column-screen-inset-left}
First the data is imported using the `readr` module from `tidyverse`. This creates two dataframes, `a_vs_b` and `a_vs_d`.
```{r}
#| echo: true
library(tidyverse)
a_vs_b <- read_tsv("Datasets/set_2/A_vs_B.deseq2.results.tsv")
a_vs_d <- read_tsv("Datasets/set_2/A_vs_D.deseq2.results.tsv")
```
:::

# Creating Summary Statistics 📏 {.column-screen-inset-left}

## Finding significantly up and down-regulated genes {.column-screen-inset-left}
To find the significantly up and down-regulated genes, there were two steps involved, firstly deciding on appropriate thresholds for what is considered 
up or down regulated and what is considered significant. 

The table below outlines the thresholds used.

|   Threshold     | log2FoldChange | Adjusted P Value |
|-----------------|----------------|------------------|
| Upregulated     |    >=0.5       | <0.05
| Downregulated   |    <=0.5       | <0.05
: {.hover}

Next a set of functions was created to find these genes in the datasets.
```{r}
#| echo: true
#Number of significantly upregulated genes
find_sig_upreg <- function(dataset) {
    #Function to take a dataframe of Deseq2 results and return a new dataframe
    #Calculates upregulated genes based on the threshold of log2fold change >0.5
    #and adjusted p value < 0.05
    changed <- subset(dataset, log2FoldChange >= 0.5 & padj < 0.05)
    return(changed)
}

#Number of significantly downregulated genes
find_sig_downreg <- function(dataset) {
    #Function to take a dataframe of Deseq2 results and return a new dataframe
    #Calculates downregulated genes based on the threshold of log2fold change <-0.5
    #and adjusted p value < 0.05
    changed <- subset(dataset, log2FoldChange <= -0.5 & padj < 0.05)
    return(changed)
}
```
This can then be ran on the two datasets to get two new dataframes of only the genes with a statistically significant (<0.05 p value) log2FoldChange.

```{r}
#| echo: true
library(DT)
sig_a_b_up <- find_sig_upreg(a_vs_b)
sig_a_b_down <- find_sig_downreg(a_vs_b)
sig_a_d_up <- find_sig_upreg(a_vs_d)
sig_a_d_down <- find_sig_downreg(a_vs_d)
```
From these dataframes, we can find the number of siginificantly changed genes:
```{r}
#| echo: true
#| code-fold: true
library(knitr)
up_down <- data.frame(Dataset_Name=c('A vs B','A vs D'),
    Upregulated_genes=c(nrow(sig_a_b_up), nrow(sig_a_d_up)),
    Downregulated_genes = c(nrow(sig_a_b_down), nrow(sig_a_d_down)))
kable(up_down)
```

## Creating a summary table {.column-screen-inset-left}
Next we need to create a summary table of the p values and log fold changes. The following function was used to generate the summary statistics for each column and dataset wanted.
```{r}
#| echo: true
#| code-fold: false
generate_summary <- function(dataset,column) {
    summary <- c(
        min(dataset[[column]],na.rm = TRUE),
        max(dataset[[column]],na.rm = TRUE),
        mean(dataset[[column]],na.rm = TRUE),
        median(dataset[[column]],na.rm = TRUE),
        quantile(dataset[[column]],0.25, na.rm = TRUE),
        quantile(dataset[[column]],0.75, na.rm = TRUE)
    )
    return(summary)
}
```
This can be used to create a dataframe for the `pvalue` and `Log2FoldChange` columns for each dataset.
```{r}
#| echo: true
stat_names = c('Min','Max','Mean','Median','Lower Quartile', 'Upper Quartile')
p_value <- generate_summary(a_vs_b,'pvalue')
log2fold <- generate_summary(a_vs_b,'log2FoldChange')
output_a_b <- data.frame(stat_names,p_value,log2fold)
kable(output_a_b, col.names=c("Statistic","P Value","Log2FoldChange"),caption = "Table 1: Summary statistics on the A vs B Dataset")
p_value <- generate_summary(a_vs_d,'pvalue')
log2fold <- generate_summary(a_vs_d,'log2FoldChange')
output_a_d <- data.frame(stat_names,p_value,log2fold)
kable(output_a_d, col.names=c("Statistic","P Value","Log2FoldChange"),caption = "Table 2: Summary statistics on the A vs D Dataset")
```


# Plotting the data 📉 {.column-screen-inset}

::: {.column-screen-inset-left}
Next was to create plots of the data including, Volcano plots, MA plots, histograms and heatmaps.
:::

::: {.callout-warning .column-body-outset-left}
**Please ensure the page is maximised before viewing graphs as they are placed left and right for easier comparisons.**
:::
All the graphs below are interactive through `plotly`.

## Volcano Plots 🌋 {.column-screen-inset}
Here are the volcano plots for the two datasets.

::: {layout-ncol=2 .quarto-full-width}
```{r}
#| echo: true
#| code-fold: true
#| label: fig-a_vs_b_volcano
#| fig-cap: A volcano plot showing the log2fold changes and up or down-regulation of genes in A vs B
library(plotly)
#Creating a new column for colouring in A vs B
a_vs_b$diffexpressed <- "NO"
a_vs_b$diffexpressed[a_vs_b$log2FoldChange >= 0.5 & a_vs_b$pvalue < 0.05] <- "UP"
a_vs_b$diffexpressed[a_vs_b$log2FoldChange <= -0.5 & a_vs_b$pvalue < 0.05] <- "DOWN"

#Creating the volcano plot (A vs B)
a_vs_b_volcano <- ggplot(data = a_vs_b, aes(x = log2FoldChange, y = -log10(pvalue), color = diffexpressed, text = gene_id)) +
    geom_point() +
    scale_color_manual(values = c("red", "grey", "blue"), labels = c('Downregulated', 'Not Significant', 'Upregulated')) +
    coord_cartesian(ylim = c(0, 20), xlim = c(-5, 5)) +
    labs(
        x = "Log 2 Fold Change",
        y = "-log10(pvalue)",
        title = "Volcano plot of A vs B",
        color = "Regulation"
    ) +
    theme_light() +
    theme(plot.title = element_text(hjust = 0.5))
ggplotly(a_vs_b_volcano)
```

```{r}
#| echo: true
#| code-fold: true
#| label: fig-a_vs_d_volcano
#| fig-cap: A volcano plot showing the log2fold changes and up or down-regulation of genes in A vs D
#Creating a new column for colouring in A vs D
a_vs_d$diffexpressed <- "NO"
a_vs_d$diffexpressed[a_vs_d$log2FoldChange >= 0.5 & a_vs_d$pvalue < 0.05] <- "UP"
a_vs_d$diffexpressed[a_vs_d$log2FoldChange <= -0.5 & a_vs_d$pvalue < 0.05] <- "DOWN"

#Creating the volcano plot (A vs D)
a_vs_d_volcano <- ggplot(data = a_vs_d, aes(x = log2FoldChange, y = -log10(pvalue), color = diffexpressed, text = gene_id)) +
    geom_point() +
    scale_color_manual(values = c("red", "grey", "blue"), labels = c('Downregulated', 'Not Significant', 'Upregulated')) +
    coord_cartesian(ylim = c(0, 40), xlim = c(-10, 10)) +
    labs(
        x = "Log 2 Fold Change",
        y = "-log10(pvalue)",
        title = "Volcano plot of A vs D",
        color = "Significance"
    ) +
    theme_light() +
    theme(plot.title = element_text(hjust = 0.5))
ggplotly(a_vs_d_volcano)
```
:::

::: {.column-screen-inset-left}
@fig-a_vs_b_volcano shows a large number of genes with down and up regulation, with not a lot of spread in the points.

@fig-a_vs_d_volcano on the other hand has a very large spread of points and with higher -log10(pvalue) scores, showing higher statistical 
significance of the change.
:::

## MA Plots 📉 {.column-screen-inset}

::: {.column-screen-inset}

MA plots were created of the two datasets using the `ggpubr` module. The plots can be used to interpret the differential expression and the expression level of the genes in A vs B and A vs D.

:::

::: {layout-ncol=2 .column-screen-inset}
```{r}
#| echo: true
#| code-fold: true
#| label: fig-a_vs_b_ma
#| fig-cap: A MA plot showing the log2fold changes and Log2 Mean Expression of A vs B
library(ggpubr)

a_vs_b_MA <- ggmaplot(data = a_vs_b
    )+
coord_cartesian(ylim = c(-10, 10)) +
labs(
    x = "Log2 Mean Expression",
    y = "Log 2 Fold Change",
    title = "MA plot of A vs B",
    color = "Significance"
) +
theme_light() +
theme(plot.title = element_text(hjust = 0.5))
ggplotly(a_vs_b_MA)
```

```{r}
#| echo: true
#| code-fold: true
#| label: fig-a_vs_d_ma
#| fig-cap: A MA plot showing the log2fold changes and Log2 Mean Expression of A vs D
a_vs_d_MA <- ggmaplot(data = a_vs_d
    )+
coord_cartesian(ylim = c(-10, 10)) +
labs(
    x = "Log2 Mean Expression",
    y = "Log 2 Fold Change",
    title = "MA plot of A vs D",
    color = "Significance"
) +
theme_light() +
theme(plot.title = element_text(hjust = 0.5))
ggplotly(a_vs_d_MA)
```

:::

Commentary of @fig-a_vs_d_ma MA plot here...


## Histograms 📊 {.column-screen-inset}
Histograms of the `pvalue` were created for each of the datasets. These along with the code, can be seen below.

::: {layout-ncol=2 .column-screen-inset}
```{r}
#| echo: true
#| code-fold: true
#Histogram of A vs B
a_vs_b_histogram <- ggplot(aes(x = pvalue), data = a_vs_b) + 
    geom_histogram(
        binwidth = 0.1,
        colour = "black",
        fill = "blue"
    ) +
    labs(
        x = "P value",
        y = "Frequency",
        title = "Histogram of P value in A vs B"  
    ) + 
    theme_light() + 
    theme(plot.title = element_text(hjust = 0.5))
ggplotly(a_vs_b_histogram)

#Histogram of A vs D
a_vs_d_histogram <- ggplot(aes(x = pvalue), data = a_vs_d) + 
    geom_histogram(
        binwidth = 0.1,
        colour = "black",
        fill = "blue"
    ) +
    labs(
        x = "P value",
        y = "Frequency",
        title = "Histogram of P value in A vs D"  
    ) + 
    theme_light() + 
    theme(plot.title = element_text(hjust = 0.5))
ggplotly(a_vs_d_histogram)
```
:::