---
title: "Gene Expression Dataset Output 🧬"
format:
    html:
        toc: true
        toc-title: "Contents"
        embed-resources: true
        page-layout: full
        theme: simplex
        code-tools: false
        code-line-numbers: true
knitr:
  opts_chunk:
    collapse: true
    message: false
    warning: false

---
This document contains all the results from and code used to complete the 'Gene Expression' portion of the LIFE4138 Coursework.

# Importing the data 📊
First the data is imported using the readr module from tidyverse. This creates two dataframes, `a_vs_b` and `a_vs_d`.
```{r}
#| echo: true
library(tidyverse)
a_vs_b <- read_tsv("Datasets/set_2/A_vs_B.deseq2.results.tsv")
a_vs_d <- read_tsv("Datasets/set_2/A_vs_D.deseq2.results.tsv")
```

# Creating Summary Statistics 📏

## Finding significantly up and down-regulated genes
To get a new dataframe of all the significantly upregulated and downregulated genes, the following functions were used:
```{r}
#| echo: true
#Number of significantly upregulated genes
find_sig_upreg <- function(dataset) {
    changed <- subset(dataset, log2FoldChange >= 1 & padj <= 0.05)
    return(changed)
}

#Number of significantly downregulated genes
find_sig_downreg <- function(dataset) {
    changed <- subset(dataset, log2FoldChange <= -1 & padj <= 0.05)
    return(changed)
}
```
This can then be ran on the two datasets to get two new dataframes of only the genes with a statistically significant (<0.05 p value) log2FoldChange.

```{r}
#| echo: true
library(DT)
sig_a_b_up <- find_sig_upreg(a_vs_b)
sig_a_b_down <- find_sig_downreg(a_vs_b)
sig_a_d_up <- find_sig_upreg(a_vs_d)
sig_a_d_down <- find_sig_downreg(a_vs_d)
```
From these dataframes, we can find the number of siginificantly changed genes:
```{r}
#| echo: true
#| code-fold: true
library(knitr)
up_down <- data.frame(Dataset_Name=c('A vs B','A vs D'),
    Upregulated_genes=c(nrow(sig_a_b_up), nrow(sig_a_d_up)),
    Downregulated_genes = c(nrow(sig_a_b_down), nrow(sig_a_d_down)))
kable(up_down)
```

Next we need to create a summary table of the p values and log fold changes.
```{r}
#| echo: true
#| code-fold: true
stat_names = c('Min','Max','Mean','Median','Lower Quartile', 'Upper Quartile')
p_value <- c(
    min(a_vs_b[["pvalue"]],na.rm = TRUE),
    max(a_vs_b[["pvalue"]],na.rm = TRUE),
    mean(a_vs_b[["pvalue"]],na.rm = TRUE),
    median(a_vs_b[["pvalue"]],na.rm = TRUE),
    quantile(a_vs_b[["pvalue"]],0.25, na.rm = TRUE),
    quantile(a_vs_b[["pvalue"]],0.75, na.rm = TRUE)
    )
log2fold <- c(
    min(a_vs_b[["log2FoldChange"]],na.rm = TRUE),
    max(a_vs_b[["log2FoldChange"]],na.rm = TRUE),
    mean(a_vs_b[["log2FoldChange"]],na.rm = TRUE),
    median(a_vs_b[["log2FoldChange"]],na.rm = TRUE),
    quantile(a_vs_b[["log2FoldChange"]],0.25, na.rm = TRUE),
    quantile(a_vs_b[["log2FoldChange"]],0.75, na.rm = TRUE)
    )
output_a_b <- data.frame(stat_names, p_value, log2fold)
kable(output_a_b)
```