---
title: "Gene Expression Dataset Output 🧬"
description: "This document contains all the results from and code used to complete the 'Gene Expression' portion of the LIFE4138 Coursework."
author: "Tahir Ansari"
date: "17 November 2025"
format:
    html:
        toc: true
        toc-title: "Contents"
        embed-resources: true
        page-layout: full
        theme: united
        code-tools: false
        code-line-numbers: true
        title-block-banner: "#4A7AB5"   
knitr:
  opts_chunk:
    collapse: true
    message: false
    warning: false
---

```{=html}
<style>
  body { background-color: #AFC4DE; }
</style>
```

# Importing the data 📊 {.column-screen-inset-left #importing-data}
::: {.column-screen-inset-left}
First the data is imported using the `readr` module from `tidyverse`. This creates two dataframes, `a_vs_b` and `a_vs_d`.
```{r}
#| echo: true
library(tidyverse)
a_vs_b <- read_tsv("Datasets/set_2/A_vs_B.deseq2.results.tsv")
a_vs_d <- read_tsv("Datasets/set_2/A_vs_D.deseq2.results.tsv")
```
:::

# Creating Summary Statistics 📏 {.column-screen-inset-left #summary-statistics}

## Finding significantly up and down-regulated genes {.column-screen-inset-left #finding-up-down-reg}
To find the significantly up and down-regulated genes, there were two steps involved, firstly deciding on appropriate thresholds for what is considered 
up or down regulated and what is considered significant. 

The table below outlines the thresholds used.

|   Threshold     | log2FoldChange | Adjusted P Value |
|-----------------|----------------|------------------|
| Upregulated     |    >=1       | <0.05
| Downregulated   |    <=1       | <0.05
: {.hover}

Next a set of functions was created to find these genes in the datasets.
```{r}
#| echo: true
#Number of significantly upregulated genes
find_sig_upreg <- function(dataset) {
    #Function to take a dataframe of Deseq2 results and return a new dataframe
    #Calculates upregulated genes based on the threshold of log2fold change >1
    #and adjusted p value < 0.05
    changed <- subset(dataset, log2FoldChange >= 1 & padj < 0.05)
    return(changed)
}

#Number of significantly downregulated genes
find_sig_downreg <- function(dataset) {
    #Function to take a dataframe of Deseq2 results and return a new dataframe
    #Calculates downregulated genes based on the threshold of log2fold change <-1
    #and adjusted p value < 0.05
    changed <- subset(dataset, log2FoldChange <= -1 & padj < 0.05)
    return(changed)
}

a_vs_b$padj[is.na(a_vs_b$padj)] <- 1 #Replaces all NAs in P adjusted value with 1 to show no significance
a_vs_d$padj[is.na(a_vs_b$padj)] <- 1 #Replaces all NAs in P adjusted value with 1 to show no significance
```
This can then be ran on the two datasets to get two new dataframes of only the genes with a statistically significant (<0.05 p value) log2FoldChange.

```{r}
#| echo: true
library(DT)
sig_a_b_up <- find_sig_upreg(a_vs_b)
sig_a_b_down <- find_sig_downreg(a_vs_b)
sig_a_d_up <- find_sig_upreg(a_vs_d)
sig_a_d_down <- find_sig_downreg(a_vs_d)
```
From these dataframes, we can find the number of siginificantly changed genes:
```{r}
#| echo: true
#| code-fold: true
library(knitr) #Loads knitr for table displaying
up_down <- data.frame(Dataset_Name=c('A vs B','A vs D'),
    Upregulated_genes=c(nrow(sig_a_b_up), nrow(sig_a_d_up)), #Num of upregulated
    Downregulated_genes = c(nrow(sig_a_b_down), nrow(sig_a_d_down))) #Num of downregulated
kable(up_down) #Shows table
```

## Creating a summary table {.column-screen-inset-left #creating-summary-table}
Next we need to create a summary table of the p values and log fold changes. The following function was used to generate the summary statistics for each column and dataset wanted.
```{r}
#| echo: true
#| code-fold: false
generate_summary <- function(dataset,column) {
    summary <- c( #Creates vector of values
        min(dataset[[column]],na.rm = TRUE), #Min value
        max(dataset[[column]],na.rm = TRUE), #Max value
        mean(dataset[[column]],na.rm = TRUE), #Mean
        median(dataset[[column]],na.rm = TRUE), #Median
        quantile(dataset[[column]],0.25, na.rm = TRUE), #First Quartile
        quantile(dataset[[column]],0.75, na.rm = TRUE) #Third Quartile
    )
    return(summary) #Returns vector of stats
}
```
This can be used to create a dataframe for the `pvalue` and `Log2FoldChange` columns for each dataset.
```{r}
#| echo: true
stat_names = c('Min','Max','Mean','Median','Lower Quartile', 'Upper Quartile') #Creates vector of stat names for table
p_value <- generate_summary(a_vs_b,'pvalue') #Generates the summary stats on the p value column
log2fold <- generate_summary(a_vs_b,'log2FoldChange') #Generates the summary stats on the log2fold change column
output_a_b <- data.frame(stat_names,p_value,log2fold) #Creates a dataframe for outputting
kable(output_a_b, col.names=c("Statistic","P Value","Log2FoldChange"),caption = "Table 1: Summary statistics on the A vs B Dataset") #Shows Table
p_value <- generate_summary(a_vs_d,'pvalue') #Generates stats for P value
log2fold <- generate_summary(a_vs_d,'log2FoldChange') #Generates stats for log2foldchange
output_a_d <- data.frame(stat_names,p_value,log2fold) #Creates dataframe for outputting
kable(output_a_d, col.names=c("Statistic","P Value","Log2FoldChange"),caption = "Table 1: Summary statistics on the A vs D Dataset") #Shows table
```

::: {.callout-warning .column-body-outset-left}
**Please ensure the page is maximised before viewing the next section as they are placed left and right for easier comparisons.**
:::

# Plotting the data 📉 {.column-screen-inset #plotting-data}

::: {.column-screen-inset-left}
Next was to create plots of the data including, Volcano plots, MA plots, histograms and heatmaps.
:::

All the graphs below are interactive through `plotly`.

## Volcano Plots 🌋 {.column-screen-inset #volcano-plots}
Here are the volcano plots for the two datasets. A function was used to generate each plot which can be seen in the fold below.

::: {layout-ncol=2 .quarto-full-width}
```{r}
#| echo: true
#| code-fold: true
#| label: fig-a_vs_b_volcano
#| code-summary: "Show the volcano plot code"
#| fig-cap: A volcano plot showing the log2fold changes against -log10(pvalue) for the genes in A vs B. Up regulated genes are shown in red (log2FC ≥ 1, padj < 0.05), down regulated genes are shown in blue (log2FC ≤ –1, padj < 0.05), while non-significant genes are shown in grey. Plots further from the centre and further from the x axis show greater significance and stronger expression.
library(plotly)
library(ggrepel)

#Creating a new column for colouring in A vs B
a_vs_b$diffexpressed <- "NO" #Creates new column and assigns NO to all values
a_vs_b$diffexpressed[a_vs_b$log2FoldChange >= 1 & a_vs_b$padj < 0.05] <- "UP" #Calculates if sig. up reg. then assigns up
a_vs_b$diffexpressed[a_vs_b$log2FoldChange <= -1 & a_vs_b$padj < 0.05] <- "DOWN" #Calculates if sig. down reg. then assigns down

#Function for volcano plot
create_volcano <- function(dataset, name, x_crop_values, y_crop_values) {
    volcano <- ggplot(data = dataset, aes(x = log2FoldChange, y = -log10(padj), color = diffexpressed, text = gene_id)) + #colours by diffexpressed and adds text to each point for plotly
        geom_point() +
        scale_color_manual(values = c("blue", "grey", "red"), labels = c('Downregulated', 'Not Significant', 'Upregulated')) + #Creates colour scheme based on labels
        coord_cartesian(ylim = y_crop_values, xlim = x_crop_values) + #Crops the graph
        labs( #Adds labels
            x = "Log 2 Fold Change",
            y = "-log10(P Adjusted Value)",
            title = paste("Volcano plot of ", name),
            color = "Significance"
        ) +
        theme_light() +
        theme(plot.title = element_text(hjust = 0.5)) #Centres title

}

#Showing the plot
ggplotly(create_volcano(a_vs_b,"A vs B",c(-5,5),c(0,15)))
```

```{r}
#| echo: true
#| code-fold: true
#| label: fig-a_vs_d_volcano
#| fig-cap: A volcano plot showing the log2fold changes against -log10(pvalue) for the genes in A vs D. Up regulated genes are shown in red (log2FC ≥ 1, padj < 0.05), down regulated genes are shown in blue (log2FC ≤ –1, padj < 0.05), while non-significant genes are shown in grey. Plots further from the centre and further from the x axis show greater significance and stronger expression.
#| code-summary: "Show the volcano plot code"
#Creating a new column for colouring in A vs D
a_vs_d$diffexpressed <- "NO" #Creates new column and assigns NO to all values
a_vs_d$diffexpressed[a_vs_d$log2FoldChange >= 1 & a_vs_d$padj < 0.05] <- "UP" #Calculates if sig. up reg. then assigns up
a_vs_d$diffexpressed[a_vs_d$log2FoldChange <= 1 & a_vs_d$padj < 0.05] <- "DOWN" #Calculates if sig. down reg. then assigns down

#Showing the plot
ggplotly(create_volcano(a_vs_d,"A vs D",c(-10,10),c(0,20)))
```
:::

::: {.column-screen-inset-left .callout}
The volcano plots show the significance and amount by which gene expression changes between conditions as well as across A vs B and A vs D. Both plots show that many genes are centred near a change of zero, with large sets showing up or down regulation.

@fig-a_vs_b_volcano shows a large number of genes with down and up regulation, with not a lot of spread in the points, while
@fig-a_vs_d_volcano shows a greater number of positive and negative changes in expression, with a larger number of significant points compared to A vs B with a much bigger spread.
:::

## MA Plots 📉 {.column-screen-inset #ma-plots}

::: {.column-screen-inset}

MA plots were created of the two datasets using the `ggpubr` module. The plots can be used to interpret the differential expression and the expression level of the genes in A vs B and A vs D.

:::

::: {layout-ncol=2 .column-screen-inset}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the MA plot code"
#| label: fig-a_vs_b_ma
#| fig-cap: MA plot of the A vs B comparison, showing Log2 Fold Change against the Log2 Mean Expression. Plots in red indicate genes that are significantly upregulated (log2FC ≥ 1, padj < 0.05), plots in blue indicate genes that significantly downregulated (log2FC ≤ –1, padj < 0.05), while plots in grey indicate genes that are not significant.
library(ggpubr) #Loads ggpubr for creating MA plots

#Function for MA plot
create_ma <- function(dataset, name,y_crop_values) {
    ma_plot <- ggmaplot(data = dataset) + #Creates an MA plot
    coord_cartesian(ylim = y_crop_values) + #Crops graph
    labs( #Adds labels
        x = "Log2 Mean Expression",
        y = "Log 2 Fold Change",
        title = paste("MA plot of ",name),
        color = "Significance"
    ) +
    theme_light() +
    theme(plot.title = element_text(hjust = 0.5)) 
}

#Drawing the MA plots
ggplotly(create_ma(a_vs_b, "A vs B", c(-10,10)))
```

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the MA plot code"
#| label: fig-a_vs_d_ma
#| fig-cap: MA plot of the A vs D comparison, showing Log2 Fold Change against the Log2 Mean Expression. Plots in red indicate genes that are significantly upregulated (log2FC ≥ 1, padj < 0.05), plots in blue indicate genes that significantly downregulated (log2FC ≤ –1, padj < 0.05), while plots in grey indicate genes that are not significant.
ggplotly(create_ma(a_vs_d, "A vs D", c(-10,10)))
```

:::
::: {.callout}
The MA plots in @fig-a_vs_b_ma and @fig-a_vs_d_ma, show the relationships between gene expression levels and log2 fold changes. In both MA plots, most plots are near zero, meaning that they have similar expression across the two conditions.  
There are distinct sets of plots that appear beyond the dashed lines, that represent significantly up and down regulated genes. @fig-a_vs_d_ma shows a greater number of these genes when compared to @fig-a_vs_b_ma, showing greater changes in gene expression in those two conditions.
:::

## Histograms 📊 {.column-screen-inset #histograms}
Histograms of the `pvalue` were created for each of the datasets. These along with the code, can be seen below.

::: {layout-ncol=2 .column-screen-inset}
```{r}
#| echo: true
#| code-fold: true
#| label: fig-a_vs_b_histogram
#| fig-cap: "Histogram plot of p-values in the A vs B comparison. Peaking towards zero suggests a large number of significantly strong differential expression values. However, a larger distribution across the higher p-values show a large number of non-significant gene expression."
#| code-summary: "Show the histogram code"

#Function to make a histogram
create_histogram <- function(dataset, name) {
    histogram_plot <- ggplot(aes(x = pvalue), data = dataset) +  #Creates histogram
    geom_histogram(
        binwidth = 0.1,
        colour = "black",
        fill = "blue",
        na.rm = TRUE #Removing any NAs to prevent errors
    ) +
    xlim(0, 1) + #Cropping along x axis
    ylim(0,1000) + #Cropping along y axis
    labs(
        x = "P value",
        y = "Frequency",
        title = paste("Histogram of P value in ", name)  
    ) + 
    theme_light() + 
    theme(plot.title = element_text(hjust = 0.5)) #Centres title
}

ggplotly(create_histogram(a_vs_b, "A vs B"))
```
```{r}
#| echo: true
#| code-fold: true
#| label: fig-a_vs_d_histogram
#| fig-cap: "Histogram plot of p-values in the A vs D comparison. Peaking towards zero is suggestive of large quantities of significantly strong differential expression values. There is a larger distribution across the higher p-values show a large number of non-significant gene expression."
#| code-summary: "Show the histogram code"
ggplotly(create_histogram(a_vs_d, "A vs D")) #Calls function to draw graph
```
:::
::: {.callout}
The histograms in @fig-a_vs_b_histogram and @fig-a_vs_d_histogram show the distribution of p values in each comparison. Both figures show that there is a large proportion of genes that show significant changes in expression because of the peak in frequency near zero.  
The overall distribution with large number of values near zero and many distributed across the rest of the range shows that many select genes were found to be important with lots showing no significance.  
DESeq2 results should show this pattern with a large number of values at zero and a uniform distribution across the rest of the range, showing that the results are what to be expected.
:::

## Heatmap ♨️ {.column-screen-inset-left #heatmap}
A heatmap was made for the top 20 differentially expressed genes from each dataset (A vs B and A vs D).

The genes with the lowest adjusted p values were selected. 

These genes had their `baseMean` value log transformed for comparison and the code and the heatmap can be seen below.

::: {layout-ncol=2 .column-screen-inset}
```{r}
#| echo: true
#| code-fold: true
#| label: fig-heatmap
#| fig-cap: Heatmap showing log10(baseMean+1) expression values for the top 20 differentially expressed genes from each dataset. Genes with the lowest adjusted p values were chosen and missing baseMean values were filled with zero values. Darker, "more blue" colours represent higher expression levels, while lighter, "more white" colours represent lower expression levels.
#| column: screen-inset-left
#| code-summary: "Show the heatmap code"
#Heatmap
heat_ab <- a_vs_b %>%
    arrange(padj) %>%
    slice_head(n = 20) %>% #Arrange by adjusted p value and take only top 20
    select(gene_id,log2FoldChange) %>%
    mutate(Name = "A_vs_B") #Add column called Name to kno which dataset its from

heat_ad <- a_vs_d %>%
    arrange(padj) %>%
    slice_head(n = 20) %>% #Arrange by adjusted p value and take only top 20
    select(gene_id,log2FoldChange) %>%
    mutate(Name = "A_vs_D") #Add column called Name to kno which dataset its from

heatmap_data <- bind_rows(heat_ab, heat_ad) #Combine the two top 20 datasets
heatmap_data <- heatmap_data %>%
    complete(gene_id,Name, fill = list(log2FoldChange = 0)) %>% #Fill blanks with 0 values
    mutate(log_expression = log2FoldChange) #Create new column with log adjusted values


heatmap_plot <- ggplot(heatmap_data, aes( #Creates the plot
    x = Name, #Comparison of datasets
    y = gene_id, #Gene list on y axis
    fill = log_expression #Colours heatmap based on expression
)) +
geom_tile() +
scale_fill_gradient(low="white", high="blue", name = "Log10(log2FoldChange)") + #Chooses colours for map
labs(
    title = "Heatmap of expression of top 20 differentially expressed genes for A vs B and A vs D",
    x = "Dataset Comparison",
    y = "Gene ID"
)+
theme_minimal() +
theme(
    panel.grid = element_blank(), #Removes gridlines
    plot.title = element_text(hjust = 0.5, size = 11) #Centres title and makes it fit
)

ggplotly(heatmap_plot)
```

::: {.callout}
@fig-heatmap shows the expression pattern for the most significantly expressed genes across the two datasets.

Many genes can be seen to show high expression in one set with low expression in the other (e.g. high in A vs B but low in A vs D). Many genes show this pattern with generally low expression in one and not the other.  

An exception to this is `Q0055` which showed high levels of expression in both sets.
:::
:::

# Significant Gene Tables 🧬 {.column-screen-inset-left #gene-tables}

Tables were generated to showcase the genes with significant differential expression for each of the two datasets. Significant genes were those that fit the same significance criteria outlined [above](#finding-up-down-reg) and can be seen once again in the table below. The same variables from the [summary table](#finding-up-down-reg) were used here.

|   Threshold     | log2FoldChange | Adjusted P Value |
|-----------------|----------------|------------------|
| Upregulated     |    >=1       | <0.05
| Downregulated   |    <=1       | <0.05
: {.hover}

Seperate tables were created for each dataset and they can be seen below.

::: {.callout-warning .column-body-outset-left}
**Only the top 25 of upregulated and downregulated genes are shown as the dataset is too large, the full list of genes is available in the files: `Sig_A_vs_B.csv` and `Sig_A_vs_D.csv` which are located in the `Significants` folder in the directory**
:::

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the significance table code"

#Creating saved full list of A vs B
sig_a_b <- bind_rows(
    sig_a_b_up %>% mutate(Regulation = "Upregulated"), #Adds regulation label
    sig_a_b_down %>% mutate(Regulation = "Downregulated") #Adds regulation label
) %>%
    select(gene_id, log2FoldChange, pvalue, padj, Regulation) #Gets only columns needed for output

dir.create(file.path("Significants"), showWarnings = FALSE) #Creates a folder to keep files tidy

write_csv(sig_a_b, "Significants/Sig_A_vs_B.csv") #Saves full list to file.

#Significance Table A vs B
sig_a_b <- bind_rows( #Creates dataframe of combined top 25
    sig_a_b_up %>%
        arrange(padj) %>% #Sort by lowest padj
        slice_head(n = 25) %>% #Take top 25
        mutate(Regulation = "Upregulated"), #Add new column to show regulation

    sig_a_b_down %>%
        arrange(padj) %>% #Sort by lowest padj
        slice_head(n = 25) %>% #Take top 25
        mutate(Regulation = "Downregulated") #Add new column to show regulation
) %>%
    select(gene_id, log2FoldChange, pvalue, padj, Regulation) #Gets all columns needed for output

datatable(sig_a_b, rownames = FALSE, caption = "Table 3: Top 25 upregulated and top 25 downregulated genes in A vs B") #Shows output nicely
```
And the table for A vs D is below.
```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the significance table code"

#Creating saved full list of A vs D
sig_a_d <- bind_rows(
    sig_a_d_up %>% mutate(Regulation = "Upregulated"), #Adds regulation label
    sig_a_d_down %>% mutate(Regulation = "Downregulated") #Adds regulation label
) %>%
    select(gene_id, log2FoldChange, pvalue, padj, Regulation) #Gets only columns needed for output

dir.create(file.path("Significants"), showWarnings = FALSE) #Creates folder to keep files tidy

write_csv(sig_a_d, "Significants/Sig_A_vs_D.csv") #Saves full list to file.

#Significance Table A vs D
sig_a_d <- bind_rows( #Creates dataframe of combined top 25
    sig_a_d_up %>%
        arrange(padj) %>% #Sort by lowest padj
        slice_head(n = 25) %>% #Take top 25
        mutate(Regulation = "Upregulated"), #Add new column to show regulation

    sig_a_d_down %>%
        arrange(padj) %>% #Sort by lowest padj
        slice_head(n = 25) %>% #Take top 25
        mutate(Regulation = "Downregulated") #Add new column to show regulation
) %>%
    select(gene_id, log2FoldChange, pvalue, padj, Regulation) #Gets all columns needed for output

datatable(sig_a_d, rownames = FALSE, caption = "Table 4: Top 25 upregulated and top 25 downregulated genes in A vs D") #Shows output nicely
```