---
title: "Gene Expression Dataset Output 🧬"
format:
    html:
        toc: true
        toc-title: "Contents"
        embed-resources: true
        page-layout: full
        theme: simplex
        code-tools: false
        code-line-numbers: true
knitr:
  opts_chunk:
    collapse: true
    message: false
    warning: false

---
This document contains all the results from and code used to complete the 'Gene Expression' portion of the LIFE4138 Coursework.

# Importing the data 📊
First the data is imported using the readr module from tidyverse. This creates two dataframes, `a_vs_b` and `a_vs_d`.
```{r}
#| echo: true
library(tidyverse)
a_vs_b <- read_tsv("Datasets/set_2/A_vs_B.deseq2.results.tsv")
a_vs_d <- read_tsv("Datasets/set_2/A_vs_D.deseq2.results.tsv")
```

# Creating Summary Statistics 📏

## Finding significantly up and down-regulated genes
To get a new dataframe of all the significantly upregulated and downregulated genes, the following functions were used:
```{r}
#| echo: true
#Number of significantly upregulated genes
find_sig_upreg <- function(dataset) {
    changed <- subset(dataset, log2FoldChange >= 1 & padj <= 0.05)
    return(changed)
}

#Number of significantly downregulated genes
find_sig_downreg <- function(dataset) {
    changed <- subset(dataset, log2FoldChange <= -1 & padj <= 0.05)
    return(changed)
}
```
This can then be ran on the two datasets to get two new dataframes of only the genes with a statistically significant (<0.05 p value) log2FoldChange.

```{r}
#| echo: true
library(DT)
sig_a_b_up <- find_sig_upreg(a_vs_b)
sig_a_b_down <- find_sig_downreg(a_vs_b)
sig_a_d_up <- find_sig_upreg(a_vs_d)
sig_a_d_down <- find_sig_downreg(a_vs_d)
```
From these dataframes, we can find the number of siginificantly changed genes:
```{r}
#| echo: true
#| code-fold: true
library(knitr)
up_down <- data.frame(Dataset_Name=c('A vs B','A vs D'),
    Upregulated_genes=c(nrow(sig_a_b_up), nrow(sig_a_d_up)),
    Downregulated_genes = c(nrow(sig_a_b_down), nrow(sig_a_d_down)))
kable(up_down)
```

## Creating a summary table
Next we need to create a summary table of the p values and log fold changes. The following function was used to generate the summary statistics for each column and dataset wanted.
```{r}
#| echo: true
#| code-fold: false
generate_summary <- function(dataset,column) {
    summary <- c(
        min(dataset[[column]],na.rm = TRUE),
        max(dataset[[column]],na.rm = TRUE),
        mean(dataset[[column]],na.rm = TRUE),
        median(dataset[[column]],na.rm = TRUE),
        quantile(dataset[[column]],0.25, na.rm = TRUE),
        quantile(dataset[[column]],0.75, na.rm = TRUE)
    )
    return(summary)
}
```
This can be used to create a dataframe for the `pvalue` and `Log2FoldChange` columns for each dataset.
```{r}
#| echo: true
stat_names = c('Min','Max','Mean','Median','Lower Quartile', 'Upper Quartile')
p_value <- generate_summary(a_vs_b,'pvalue')
log2fold <- generate_summary(a_vs_b,'log2FoldChange')
output_a_b <- data.frame(stat_names,p_value,log2fold)
kable(output_a_b, col.names=c("Statistic","P Value","Log2FoldChange"),caption = "Table 1: Summary statistics on the A vs B Dataset")
p_value <- generate_summary(a_vs_d,'pvalue')
log2fold <- generate_summary(a_vs_d,'log2FoldChange')
output_a_d <- data.frame(stat_names,p_value,log2fold)
kable(output_a_d, col.names=c("Statistic","P Value","Log2FoldChange"),caption = "Table 2: Summary statistics on the A vs D Dataset")
```

# Volcano Plots 🌋
Here are the volcano plots for the two datasets.

```{r}
#| echo: true
#| code-fold: true
#| label: fig-a_vs_b_volcano
#| fig-cap: A volcano plot showing the log2fold changes and up or down-regulation of genes in A vs B
library(plotly)
#Creating a new column for colouring in A vs B
a_vs_b$diffexpressed <- "NO"
a_vs_b$diffexpressed[a_vs_b$log2FoldChange >= 0.5 & a_vs_b$pvalue < 0.05] <- "UP"
a_vs_b$diffexpressed[a_vs_b$log2FoldChange <= -0.5 & a_vs_b$pvalue < 0.05] <- "DOWN"

#Creating the volcano plot (A vs B)
a_vs_b_volcano <- ggplot(data = a_vs_b, aes(x = log2FoldChange, y = -log10(pvalue), color = diffexpressed, text = gene_id)) +
    geom_point() +
    scale_color_manual(values = c("red", "grey", "blue"), labels = c('Downregulated', 'Not Significant', 'Upregulated')) +
    coord_cartesian(ylim = c(0, 20), xlim = c(-5, 5)) +
    labs(
        x = "Log 2 Fold Change",
        y = "-log10(pvalue)",
        title = "Volcano plot of A vs B",
        color = "Regulation"
    ) +
    theme_light() +
    theme(plot.title = element_text(hjust = 0.5))
ggplotly(a_vs_b_volcano)
```
```{r}
#| echo: true
#| code-fold: true
#| label: fig-a_vs_d_volcano
#| fig-cap: A volcano plot showing the log2fold changes and up or down-regulation of genes in A vs D
#Creating a new column for colouring in A vs D
a_vs_d$diffexpressed <- "NO"
a_vs_d$diffexpressed[a_vs_d$log2FoldChange >= 0.5 & a_vs_d$pvalue < 0.05] <- "UP"
a_vs_d$diffexpressed[a_vs_d$log2FoldChange <= -0.5 & a_vs_d$pvalue < 0.05] <- "DOWN"

#Creating the volcano plot (A vs B)
a_vs_d_volcano <- ggplot(data = a_vs_d, aes(x = log2FoldChange, y = -log10(pvalue), color = diffexpressed, text = gene_id)) +
    geom_point() +
    scale_color_manual(values = c("red", "grey", "blue"), labels = c('Downregulated', 'Not Significant', 'Upregulated')) +
    coord_cartesian(ylim = c(0, 40), xlim = c(-10, 10)) +
    labs(
        x = "Log 2 Fold Change",
        y = "-log10(pvalue)",
        title = "Volcano plot of A vs D",
        color = "Significance"
    ) +
    theme_light() +
    theme(plot.title = element_text(hjust = 0.5))
ggplotly(a_vs_d_volcano)
```